Notes and other text

### 2023-04-06 doc/manuscript/translocation.qmd
Add text and comments from Allie and Bree

Host evolution following disease-induced declines is one mechanism by which hosts can recover [@carlson2014; @searle2020].
"Evolutionary rescue" describes rapid evolutionary change originating from variation within a host population that reverses ongoing population declines [@carlson2014] <!--# Bree comment: Nuances of word choice -->.
Evolutionary changes following disease-induced declines are known from a diversity of wildlife populations [@savage2016; @epstein2016; @gignoux-wolfsohn2021; @holland2022], but linking these selective responses directly to population recoveries, as necessary for evolutionary rescue, is difficult and remains rare.
In addition to promoting recovery of remnant populations after disease-induced declines, selection that promotes recovery of focal populations can also provide sources of individuals with increased resistance and/or tolerance (hereafter, "resistance") to reestablish populations extirpated following disease emergence [e.g., @mendelson2019].
To date, the use of resistant individuals for species recovery is rare [but see @joseph2018], and studies that evaluate its potential utility are critically needed.
<!--# Other examples in addition to Joseph and Knapp (2018)? In other taxa besides amphibians? -->

There are four conditions that will often determine the feasibility of using individuals from populations that are recovering following disease-induced declines to reestablish extirpated populations.
First, populations that survived following disease arrival have evolved increased resistance.
Second, increased resistance contributes to higher individual fitness in the presence of disease <!--# "disease" or "the pathogen"? Resistance could convey complete prevention of disease, so I often struggle with which word to use -->and allows positive population growth.
Third, the fitness benefits gained through selection for increased resistance manifest beyond the natal environment and are maintained across generations, allowing for the establishment of reintroduced populations.
And fourth, reintroduced populations show long-term growth and persistence.
To date, few of these conditions are documented for any single species impacted by disease, and as a result the reintroduction of resistant hosts as a means to recover species of conservation concern remains largely an untested idea [@scheele2021; @mendelson2019].

The mountain yellow-legged (MYL) frog, comprised of the sister species *Rana muscosa* and *Rana sierrae* [@vredenburg2007], is emblematic of the unprecedented global declines of amphibians caused by the recently-emerged amphibian chytrid fungus [*Batrachochytrium dendrobatidis,* @scheele2019].
Once the most common amphibian in the high elevation portion of the Sierra Nevada mountains [California, USA, @grinnell1924], this frog has disappeared from more than 90% of its historical range during the past century [@vredenburg2007].
Due to the severity of its decline and the increasing probability of extinction, both species are now listed as "endangered" under the U.S.
Endangered Species Act.
In the Sierra Nevada, this decline was initiated by the introduction of non-native trout into the extensive historically-fishless region [@bradford1989; @knapp2000], and the arrival of *B. dendrobatidis* (Bd) in the 1960s and its subsequent spread [@vredenburg2019] caused additional widespread population extirpations [@vredenburg2010; @rachowicz2006].
These Bd-caused declines are fundamentally different from the fish-caused declines because removal of fish populations is feasible and results in the rapid recovery of frogs [@knapp2007; @vredenburg2004].
In contrast, Bd appears to persist in habitats even in the absence of amphibian hosts [@walker2007], and therefore represents a permanent alteration of invaded ecosystems that amphibians will need to overcome to reestablish populations.

@@

To detect regions of the genome under selection, we used three complementary approaches that together describe the evolutionary response of frog populations to Bd.
First, we used a multivariate linear mixed model to evaluate associations between population type (i.e., naive versus recovering) and individual variants.
For these single nucleotide polymorphisms (SNPs), we identified outlier variants using two Bonferroni-corrected significance levels: "liberal" = 0.05, "stringent" = 0.01.
Second, we used a splined window approach to evaluate larger genomic regions showing differentiation between population types in *F~ST~* and nucleotide diversity ($\pi$).
In this analysis, we identified outlier regions using a threshold for the W-statistic (4 standard deviations from the mean) <!--# This is a different definition than the one given in the Methods: "4 * standard deviation". -->.
Third, we assessed whether putative genes or genomic regions that showed strong evidence of selection were associated with biological functions, molecular functions, or cellular processes (i.e., GO terms) consistent with possible resistance mechanisms.

<!--# I suggest we mention the PCA in the paragraph above so it doesn't come as a surprise here. -->Based on the genomic data, individual frogs clustered into three separate groups in principal component space (@fig-selectionresults B: SI), and clusters reflected the strong signature of spatial proximity on genetic structure characteristic of MYL frogs [@rothstein2020; @poorten2017].
Importantly, each cluster contained at least one population from both the naive and recovering groups (@fig-selectionresults B: SI) <!--# Add text indicating why this is important for interpretation of the selection analyses -->.
Our first analysis to detect signatures of selection identified a liberal set of variants that included 38 outliers (35 SNPs and 3 structural variants) from 30 distinct genes across 16 contigs.
The stringent set contained 11 outlier SNPs from 7 distinct genes across 4 contigs (@fig-selectionresults C, D: SI; SI Dataset 1).
A single gene in both groups did not have an associated annotation.
For the SNPs in the stringent set of outliers, frequency differences between the naive and recovering populations ranged from 41% to 86%.
Most of these SNPs showed frequency differences in only a subset of the sampled populations (@fig-allelefrequencies A, B) <!--# Move this caveat text to the Discussion? :  and the patterns may be confounded by frog relatedness and/or population structure despite the methods we used to account for these factors. -->, but the outlier SNP in the RIN3 gene showed consistent differences in frequencies across all populations (@fig-allelefrequencies C).
This is suggestive of strong selection at this locus associated with exposure of frogs to Bd.
No GO terms were significantly over-represented in the set of 29 annotated outlier genes.

Our second analysis, using a splined window approach, identified regions of the genome that had both high differentiation (*F~ST~*) and a large difference in nucleotide diversity between naive and recovering populations ($\pi_{diff}$ = $\pi_{naive}$ - $\pi_{recovering}$).
We identified 58 outlier windows for *F~ST~* and 33 outlier windows for $\pi_{diff}$ (@fig-spline-manhattan A, B).
Of these, 9 regions were outliers for both metrics and two of these shared regions also contained outlier SNPs.<!--# "windows" and "regions" are used interchangeably in this paragraph. To avoid confusion, use "windows" throughout? --> A total of 33 annotated genes were found in the 9 shared regions (SI Dataset 2 <!--# In the supplementary_table_outliers_typofix.xlsx file > the splined_window_shared_ouliers sheet contains descriptions of 10 shared regions, not 9-->) and no GO terms were overrepresented within this set of genes.
The strongest signal for a reduction in $\pi$ in recovering populations (highest $\pi_{diff}$), indicative of directional selection, occurred in a 163kb region on Contig19, 12.9Mb upstream of the RIN3 outlier SNP (@fig-spline-manhattan C).
This region contains \~500 SNPs and one annotated gene called "interferon-induced very large GTPase 1-like" (GVINP1).
Additionally, a shared outlier region on Contig1 contained two complement factor genes (C6 and C7).
Interestingly, this region had higher $\pi$ in the recovering populations, consistent with a pattern of balancing selection.
Finally, one shared outlier region on Contig8 contained one outlier SNP (TCF19) and was within 360kb of another outlier SNP (VARS) (@fig-spline-manhattan D).
This region (854kb from the beginning of the outlier windows to the VARS SNP) contained a total of 10 annotated genes.
In *Xenopus*, four of these genes occur in the extended MHC Class I region (FLOT1, TUBB, MDC1,TCF19) and three occur in the extended MHC Class II region (HSP70, LSM2, VARS) [@ohta2006].
Therefore, we conclude that this candidate region under selection is a part of the extended MHC Class I and II complex and shows synteny with other amphibian genomes.
<!--# Here or in the discussion we will need to place this result in the context of previous work on MHC in amphibians.  Readers not familiar with amphibian-Bd systems might miss the importance of this result --><!--# Based on Fu and Waldman (2017), are any of our data sufficient to conclude anything related to statement that "Bd-resistant species share similar pocket conformations within the MHC-II antigen-binding groove. Among susceptible species, survivors of epizootics bear alleles encoding these conformations."? What about MHC Class II characteristics linked to resistance by Bataille et al. (2015)? Savage and Zamudio (2011, 2016), Kosch et al. (2018)?-->

@@

<!--# Bree comment: Here or elsewhere - it feels like we need a more specific bridge from the genomics section into the translocation section. I think we need to be explicit about how the populations we chose for the genomics section relate to the populations used in translocations. Even if they aren't matched in any way, we should reiterate something about how the experimental designs comport. For example, we could explain how the genomic analyses were conducted with multiple recovering populations so we know evidence for selection is strong in multiple areas of the genome in multiple recovering populations. -->

@@

To identify site and frog-specific drivers of post-translocation survival, we developed linear and multilevel models in a Bayesian framework.
Our specific interest was whether Bd load had a negative effect on adult survival, as would be expected if frogs were highly susceptible to Bd infection.
The best model of 1-year frog survival identified several important predictors, but Bd load at the time of translocation was not among them (bd_load; @fig-survival-postdens SI).
Important predictors included winter severity in the year following translocation (snow_t1), site elevation (elevation), and donor population (donor_72996; @fig-cond-effects).
Males had somewhat higher survival than females, but this effect was only marginally important (sex_male; @fig-cond-effects, @fig-survival-postdens SI).
In addition to Bd load, effects of frog size (length), winter severity in the year of translocation (snow_t), day of year on which a translocation was conducted (day), and translocation order (order_first) were all unimportant (@fig-survival-postdens SI).
Collectively, these results show that Bd load is not a predictor of the survival of translocated frogs, consistent with frogs in recovering populations having evolved sufficient resistance to minimize the adverse effects of Bd infection.
<!--# Allie comment: Maybe tolerance is a better word here? Since they were infected. Or maybe we need to explicitly define what we mean with resistance vs tolerance. Bree comment: I noticed this also – in the Intro, you do say that you are lumping both resistance and tolerance into the phrase "resistance" but you are talking about multiple things (being able to avoid infection, being able to fight infection, being able to tolerate infection, etc). Maybe at some point we want to address specifically that the genomic changes we are observing could contribute to multiple facets of resistance / tolerance.  And that we are lumping those together in an "evolved resistance" bin. --> In addition, the moderate Bd loads that characterized frogs following translocation (@fig-bdload-beforeafter SI) and relatively high survival of frogs in numerous translocated cohorts (@fig-translocation-survival) suggests that resistant frogs from recovering populations can maintain the benefits of resistance in non-natal habitats.
Finally, in three locations where longer CMR time series allowed us to assess the survival of adults recruited to the population (i.e., progeny of translocated individuals), naturally recruited adults had equivalent or higher survival probabilities than the originally translocated adults (@fig-compare_surv_probs SI).
Therefore, frog resistance in non-natal habitats is maintained across generations.

@@

Effectively leveraging evolutionary rescue for landscape-level species recovery requires strong evidence that evolutionary rescue is in fact the mechanism underlying recovery.
In the MYL frog system, there are four lines of evidence that support evolutionary rescue as primary mechanism of recovery.
First, in host-pathogen systems evolutionary rescue is more likely to occur when hosts experience repeated selection events (e.g., through re-infections), there is a persistent pathogen reservoir or frequency-dependent infection dynamics, and populations experience increased per capita birth rates at small population sizes [@golas2021].
At least two of these conditions are met in the MYL frog system: i) individuals that recover from Bd infection do not gain permanent immunity to Bd and are susceptible to re-infection (Knapp, PeerJ), and ii) there are potential Bd reservoirs (Briggs, Reeder) and seasonal frequency-dependent transmission dynamics (Wilber) <!--# This first line of evidence is weak and could be dropped. Just because conditions under which ER is most likely exist in MYL frogs doesn't mean much regarding whether ER is the most likely mechanism underlying recovery. -->.
Second, our genomic analyses comparing naive and recovering MYL frog populations identified notable differences in allele frequencies between adjacent Bd-naive and recovering populations.
These differences are consistent with strong selection and some regions of the genome under selection are associated with known cellular mechanisms associated with resistance, such those related amphibian immune responses <!--# Add specifics to make this stronger, broaden to include not just allele frequencies but genomic regions. -->.
Third, our translocations and long-term assessment of establishment dynamics demonstrate that evolved resistance extends beyond the natal environment and into subsequent generations, as demonstrated by high survival rates of translocated frogs and recruits into the adult population in the presence of Bd.
Finally, resistance can be statistically quantified --- in recovering populations and those established via translocation of frogs from recovering populations, Bd load is not an important predictor of post-translocation frog survival (in contrast to important relationship between Bd load and frog survival during Bd-caused epizootics) <!--# Whether this resistance is due to evolution or adaptive immunity is not distinguishable under this line of evidence. -->.
Although other studies have documented signatures of selection in hosts following disease-induced declines (Epstein, white-nose, Byrne), few have demonstrated evolutionary rescue: disease-induced selection of host resistance and subsequent host recovery.
Our study is unique in that we provide strong evidence that evolved resistance begets population recovery in both natal and non-natal habitats, consistent with evolutionary rescue.

@@

@@

@@

Raw reads were filtered for adapters and contaminants using fastp [@chen2018] and aligned to the *Rana muscosa* genome [NCBI SRA: SRS6533475, reference = @hon2020] with repetitive elements masked using bwa ("mem" mode) [@li2013].
Exact PCR Duplicates were marked using Picard.
Variants were then called following GATK best practices (v.XX, GATK ref).
Briefly, raw variants were called for each sample using HaplotypeCaller and combined using CombineGVCFs.
Next, genotypes were jointly called using GenotypeGVCFs.
Variants were then hard filtered using gatk VariantFiltration using the following parameters to remove low-quality sites: QD \< 2.0, FS \> 40.0, SOR \> 3.0, MQ \< 50.0, MQRankSum \> 3.0, MQRankSum \< -3.0, ReadPosRankSum \> 3.0, ReadPosRankSum \< -3.0.
This initial filter resulted in 1,595,206 variant sites across 53 individuals.

@@

To visualize the genomic relationship of our populations we created a PCA using the glPCA function in the adegenet R package [@jombart2008].
Next, to detect regions of the genome under selection between naive and recovering populations, we used a multivariate linear mixed model to evaluate individual variants and a splined window analysis to evaluate larger genomic regions.
<!--# Improve wording in this and subsequent paragraph to clarify that the two approaches were used to identify regions under selection.  For example, for us genomics dummies it would be useful to see something like: "We used two related approaches to identify regions under selection. The first approach A, did X, Y, and Z but had C limitations.  The second approach B, did X, Y, and Z and D limitations."-->.
GEMMA calculates and incorporates a relatedness matrix for input samples, allowing us to account for relatedness and population structure when calculating likelihood ratio statistics for our association test.
For this analysis we used a stringent data filter to include only SNPs with \< 5% missing data (missing for no more than 2 individuals).
GEMMA then calculated the likelihood ratio statistic <!--# What is the null with this test (or what are the two models being compared with the LRT)?  Probably obvious for genomics folks, but not totally clear to me.  I am guessing we are testing whether the `population` coefficient (naive vs. persistent) is significant for describing variability in a SNP? --> for the set of 148,307 high quality SNPs across 127 contigs.
We then calculated two statistical thresholds for identifying outliers using a Bonferroni correction for a significance level of 0.05 (liberal set) and 0.01 (stringent set).
<!--# Could we say something like: "Outliers indicated that a SNP was differently represented in naive vs. persistent populations" or whatever is the correct language." --> Results were visualized using a Manhattan plot and qqplot.

@@

##

>>>>

Host evolution following disease-induced declines is one mechanism by which hosts can recover [@carlson2014; @searle2020].
"Evolutionary rescue" describes rapid evolutionary change originating from variation within a host population that reverses ongoing population declines [@carlson2014] <!--# Allie comment. I believe evolutionary rescue also encompasses the introduction of new variation through gene flow. See box 1 in ref 16 for a helpful discussion of genetic vs evolutionary rescue. Could delete "Originating from variation within a host population" here and just add more information about what make evolutionary rescue special - that is the increase in locally adapted phenotypes. Bree comment. Minor – there are a lot of clauses in this sentence. Wonder if setting this as " – as necessary for evolution rescue – " might make the sentence more readable-->.
Evolutionary changes following disease-induced declines are known from a diversity of wildlife populations [@savage2016; @epstein2016; @gignoux-wolfsohn2021; @holland2022], but linking these selective responses directly to population recoveries, as necessary for evolutionary rescue, is difficult and remains rare.
In addition to promoting recovery of remnant populations after disease-induced declines, selection that promotes recovery of focal populations can also provide sources of individuals <!--# Bree comment: Again minor, but seems like this could be slimmed (as the sentence repeats the idea of promoting recovery twice and the "provide sources of individuals" seems awkward). It could read something like "In addition to promoting recovery of remnant populations after disease-induced declines, natural selection can also lead to increased resistance and/or tolerance…." -->with increased resistance and/or tolerance (hereafter, "resistance") to reestablish populations extirpated following disease emergence [e.g., @mendelson2019].
To date, the use of resistant individuals for species recovery is rare [but see @joseph2018], and studies that evaluate its potential utility are critically needed.
<!--# Other examples in addition to Joseph and Knapp (2018)? In other taxa besides amphibians? -->

There are four conditions that will often determine the feasibility of using individuals from populations that are recovering following disease-induced declines to reestablish extirpated populations.
First, populations that survived <!--# Bree comment. Worth keeping tense consistent so that the four conditions feel parallel? This sentence could read "First, populations that survive following disease arrival evolve increased resistance" -->following disease arrival have evolved increased resistance.
Second, increased resistance contributes to higher individual fitness in the presence of disease <!--# "disease" or "the pathogen"? Resistance could convey complete prevention of disease, so I often struggle with which word to use -->and allows positive population growth.
Third, the fitness benefits gained through selection for increased resistance manifest beyond the natal environment and are maintained across generations, allowing for the establishment of reintroduced populations.
And fourth, reintroduced populations show long-term growth and persistence.
To date, few of these conditions are documented for any single species impacted by disease, and as a result the reintroduction of resistant hosts as a means to recover species of conservation concern remains largely an untested idea [@scheele2021; @mendelson2019].

The mountain yellow-legged (MYL) frog, comprised of the sister species *Rana muscosa* and *Rana sierrae* [@vredenburg2007], is emblematic of the unprecedented global declines of amphibians caused by the recently-emerged amphibian chytrid fungus [*Batrachochytrium dendrobatidis,* @scheele2019].
Once the most common amphibian in the high elevation portion of the Sierra Nevada mountains [California, USA, @grinnell1924], this frog has disappeared from more than 90% of its historical range during the past century [@vredenburg2007].
Due to the severity of its decline and the increasing probability of extinction, both species are now listed as "endangered" under the U.S.
Endangered Species Act.
In the Sierra Nevada, this decline was initiated by the introduction of non-native trout into the extensive historically-fishless region [@bradford1989; @knapp2000] <!--# Bree comment: Very minor but there are lots of places with long structurally complex sentences that seem just as good as broken up shorter sentences -->, and the arrival of *B. dendrobatidis* (Bd) in the 1960s and its subsequent spread [@vredenburg2019] caused additional widespread population extirpations [@vredenburg2010; @rachowicz2006].
These Bd-caused declines are fundamentally different from the fish-caused declines because removal of fish populations is feasible and results in the rapid recovery of frogs [@knapp2007; @vredenburg2004].
In contrast, Bd appears to persist in habitats even in the absence of amphibian hosts [@walker2007], and therefore represents a long-term alteration of invaded ecosystems that amphibians will need to overcome to reestablish populations.

@@

To detect regions of the genome under selection, we used three complementary approaches that together describe the evolutionary response of frog populations to Bd.
First, we used a multivariate linear mixed model to evaluate associations between population type (i.e., naive versus recovering) and individual variants.
For these single nucleotide polymorphisms (SNPs), we identified outlier variants using two Bonferroni-corrected significance levels: "liberal" = 0.05, "stringent" = 0.01.
Second, we used a splined window approach to evaluate larger genomic regions showing differentiation between population types in *F~ST~* and nucleotide diversity ($\pi$) <!--# Allie comment: Feels like we need to signal here that we are calculating the difference in pi between naive and persistent. pi-diff.  -->.
In this analysis, we identified outlier regions using a threshold for the W-statistic (4 standard deviations from the mean) and selected regions where outlier windows overalapped for *F~ST~* and $\pi$ <!--# This is a different definition than the one given in the Methods: "4 * standard deviation". --> <!--# Allie comment: Feels like we need to signal here that we are calculating the difference in pi between naive and persistent. pi - diff -->.
Third, we assessed whether putative genes or genomic regions that showed strong evidence of selection were associated with biological functions, molecular functions, or cellular processes (i.e., GO terms) consistent with possible resistance mechanisms.

<!--# I suggest we mention the PCA in the paragraph above so it doesn't come as a surprise here. -->

Based on the genomic data, individual frogs clustered into three separate groups in principal component space (@fig-selectionresults B: SI), and clusters reflected the strong signature of spatial proximity on genetic structure characteristic of MYL frogs [@rothstein2020; @poorten2017].
Importantly, each cluster contained at least one population from both the naive and recovering groups (@fig-selectionresults B: SI) <!--# Add text indicating why this is important for interpretation of the selection analyses -->.
Our first analysis to detect signatures of selection identified a liberal set of variants that included 38 outliers (35 SNPs and 3 structural variants) from 30 distinct genes across 16 contigs.
The stringent set contained 11 outlier SNPs from 7 distinct genes across 4 contigs (@fig-selectionresults C, D: SI; SI Dataset 1).
A single gene in both groups did not have an associated annotation.
For the SNPs in the stringent set of outliers, frequency differences between the naive and recovering populations ranged from 41% to 86%.
Most of these SNPs showed frequency differences in only a subset of the sampled populations (@fig-allelefrequencies A, B) <!--# Move this caveat text to the Discussion? :  and the patterns may be confounded by frog relatedness and/or population structure despite the methods we used to account for these factors. -->, but the outlier SNP in the RIN3 gene showed consistent differences in frequencies across all populations (@fig-allelefrequencies C).
This is suggestive of strong selection at this locus associated with exposure of frogs to Bd.
No GO terms were significantly over-represented in the set of 29 annotated outlier genes.

Our second analysis, using a splined window approach, identified regions of the genome that had both high differentiation (*F~ST~*) and a large difference in nucleotide diversity between naive and recovering populations ($\pi_{diff}$ = $\pi_{naive}$ - $\pi_{recovering}$).
We identified 58 outlier windows for *F~ST~* and 33 outlier windows for $\pi_{diff}$ (@fig-spline-manhattan A, B).
Of these, 9 regions were outliers for both metrics and two of these shared regions also contained outlier SNPs.<!--# "windows" and "regions" are used interchangeably in this paragraph. To avoid confusion, use "windows" throughout? --> A total of 33 annotated genes were found in the 9 shared regions (SI Dataset 2 <!--# In the supplementary_table_outliers_typofix.xlsx file > the splined_window_shared_ouliers sheet contains descriptions of 10 shared regions, not 9--> <!--# Allie comment: One is a repeat because there is a 2>1 match for one of the overlaps (i.e., One Fst outlier window overlaps with 2 pi outlier windows).  -->) and no GO terms were overrepresented within this set of genes.
The strongest signal for a reduction in $\pi$ in recovering populations (highest $\pi_{diff}$) <!--# Allie comment: For me this is enough of a reason to discuss this beyond the obvious relevance of the gene function.  -->, indicative of directional selection, occurred in a 163kb region on Contig19, 12.9Mb upstream of the RIN3 outlier SNP (@fig-spline-manhattan C).
This region contains \~500 SNPs and one annotated gene called "interferon-induced very large GTPase 1-like" (GVINP1 <!--# Allie comment: This gene is even more interesting now that I've done a lit review. More in the discussion! -->).
Additionally, a shared outlier region on Contig1 contained two complement factor genes (C6 and C7).
Interestingly, this region had higher $\pi$ in the recovering populations, consistent with a pattern of balancing selection.
Finally, one shared outlier region on Contig8 contained one outlier SNP (TCF19) <!--# Allie comment: Again, this is a very good reason to discuss - there are both outlier windows and individuals SNPS here making it an important part outlier region to discuss BEYOND the cool MHC connection.  -->and was within 360kb of another outlier SNP (VARS) (@fig-spline-manhattan D).
This region (854kb from the beginning of the outlier windows to the VARS SNP) contained a total of 10 annotated genes.
In *Xenopus*, four of these genes occur in the extended MHC Class I region (FLOT1, TUBB, MDC1,TCF19) and three occur in the extended MHC Class II region (HSP70, LSM2, VARS) [@ohta2006].
Therefore, we conclude that this candidate region under selection is a part of the extended MHC Class I and II complex and shows synteny with other amphibian genomes.
<!--# Here or in the discussion we will need to place this result in the context of previous work on MHC in amphibians.  Readers not familiar with amphibian-Bd systems might miss the importance of this result --><!--# Based on Fu and Waldman (2017), are any of our data sufficient to conclude anything related to statement that "Bd-resistant species share similar pocket conformations within the MHC-II antigen-binding groove. Among susceptible species, survivors of epizootics bear alleles encoding these conformations."? What about MHC Class II characteristics linked to resistance by Bataille et al. (2015)? Savage and Zamudio (2011, 2016), Kosch et al. (2018)?-->

@@

<!--# Bree comment: Here or elsewhere - it feels like we need a more specific bridge from the genomics section into the translocation section. I think we need to be explicit about how the populations we chose for the genomics section relate to the populations used in translocations. Even if they aren't matched in any way, we should reiterate something about how the experimental designs comport. For example, we could explain how the genomic analyses were conducted with multiple recovering populations so we know evidence for selection is strong in multiple areas of the genome in multiple recovering populations. Roland response: I added what I hope are bridging sentences between all three sections. -->

@@

To identify site and frog-specific drivers of post-translocation survival, we developed linear and multilevel models in a Bayesian framework.
Our specific interest was whether Bd load had a negative effect on adult survival, as would be expected if frogs were highly susceptible to Bd infection.
The best model of 1-year frog survival identified several important predictors, but Bd load at the time of translocation was not among them (bd_load; @fig-survival-postdens SI).
Important predictors included winter severity in the year following translocation (snow_t1), site elevation (elevation), and donor population (donor_72996; @fig-cond-effects).
Males had somewhat higher survival than females, but this effect was only marginally important (sex_male; @fig-cond-effects, @fig-survival-postdens SI).
In addition to Bd load, effects of frog size (length), winter severity in the year of translocation (snow_t), day of year on which a translocation was conducted (day), and translocation order (order_first) were all unimportant (@fig-survival-postdens SI).
Collectively, these results show that Bd load is not a predictor of the survival of translocated frogs, consistent with frogs in recovering populations having evolved sufficient resistance to minimize the adverse effects of Bd infection.
<!--# Allie comment: Maybe tolerance is a better word here? Since they were infected. Or maybe we need to explicitly define what we mean with resistance vs tolerance. Bree comment: I noticed this also – in the Intro, you do say that you are lumping both resistance and tolerance into the phrase "resistance" but you are talking about multiple things (being able to avoid infection, being able to fight infection, being able to tolerate infection, etc). Maybe at some point we want to address specifically that the genomic changes we are observing could contribute to multiple facets of resistance / tolerance.  And that we are lumping those together in an "evolved resistance" bin. Roland comment: We reworded this sentence, but probably have not fully dealt with resistance/tolerance issue. --> In addition, the moderate Bd loads that characterized frogs following translocation (@fig-bdload-beforeafter SI) and relatively high survival of frogs in numerous translocated cohorts (@fig-translocation-survival) suggests that resistant frogs from recovering populations can maintain the benefits of resistance in non-natal habitats.
Finally, in three locations where longer CMR time series allowed us to assess the survival of adults recruited to the population (i.e., progeny of translocated individuals), naturally recruited adults had equivalent or higher survival probabilities than the originally translocated adults (@fig-compare_surv_probs SI).
Therefore, frog resistance in non-natal habitats is maintained across generations.

@@

<!--# Bree comment: It seems to me that this paragraph might be better as a shorter high level compelling summary and then the subsequent sections of the Discussion go into each of the four lines of evidence in more detail. Otherwise it feels a bit heavy and I'm not sure readers will be able to hold it all in mind at once.  -->Effectively leveraging evolutionary rescue for landscape-level species recovery requires strong evidence that evolutionary rescue is in fact the mechanism underlying recovery.
In the MYL frog system, there are four lines of evidence that support evolutionary rescue as primary mechanism of recovery.
First, in host-pathogen systems evolutionary rescue is more likely to occur when hosts experience repeated selection events (e.g., through re-infections), there is a persistent pathogen reservoir or frequency-dependent infection dynamics, and populations experience increased per capita birth rates at small population sizes [@golas2021].
At least two of these conditions are met in the MYL frog system: i) individuals that recover from Bd infection do not gain permanent immunity to Bd and are susceptible to re-infection (Knapp, PeerJ), and ii) there are potential Bd reservoirs (Briggs, Reeder) and seasonal frequency-dependent transmission dynamics (Wilber) <!--# This first line of evidence is weak and could be dropped. Just because conditions under which ER is most likely exist in MYL frogs doesn't mean much regarding whether ER is the most likely mechanism underlying recovery. -->.
Second, our genomic analyses comparing naive and recovering MYL frog populations identified notable differences in allele frequencies between adjacent Bd-naive and recovering populations.
These differences are consistent with strong selection and some regions of the genome under selection are associated with known cellular mechanisms associated with resistance, such those related amphibian immune responses <!--# Add specifics to make this stronger, broaden to include not just allele frequencies but genomic regions. -->.
Third, our translocations and long-term assessment of establishment dynamics demonstrate that evolved resistance extends beyond the natal environment and into subsequent generations, as demonstrated by high survival rates of translocated frogs and recruits into the adult population in the presence of Bd.
Finally, resistance can be statistically quantified --- in recovering populations and those established via translocation of frogs from recovering populations, Bd load is not an important predictor of post-translocation frog survival (in contrast to important relationship between Bd load and frog survival during Bd-caused epizootics) <!--# Whether this resistance is due to evolution or adaptive immunity is not distinguishable under this line of evidence. -->.
Although other studies have documented signatures of selection in hosts following disease-induced declines (Epstein, white-nose, Byrne), few have demonstrated evolutionary rescue: disease-induced selection of host resistance and subsequent host recovery.
Our study is unique in that we provide strong evidence that evolved resistance begets population recovery in both natal and non-natal habitats, consistent with evolutionary rescue.

@@

<!--# The next 2 paragraphs are new text from Allie -->Multiple genes in our set of outliers are strong candidates for playing a role the genomic underpinnings of Bd-survival in MYL frogs.
Remarkably, we identified a \~940kb region on contig 8 that contains 8 genes associated with either the MHC Class I or Class III regions.
The order of these genes in the MYL frog genome is conserved from *Xenopus* [@ohta2006] and is consistent with MHC Class I/III linkage reported in the Komodo dragon [@reed2021].
These results corroborate many studies linking MHC genes to Bd survival [e.g., @savage2011; @bataille2015].
For example, the MHC Class III gene heat shock protein 70 (HSP70) within this outlier window is intriguing because an HSP70-binding protein was also identified as a candidate gene under selection in persisting *Atelopus varius* populations [@byrne2020], and HSP70 was highly expressed in Bd-infected vs. uninfected *Xenopus tropicalis* [@rosenblum2009].
This MHC gene cluster lies on the same chromosome as the outlier SNP in RIN3, a gene associated with inflammation response that is highly expressed in amputated Xenopus tails [@fukazawa2009].
These findings suggest that genetic variants in this region of the MHC, and in other immunologically relevant genes on the same chromosome, are likely important for frog resistance against Bd.

Additionally, we identified a region on contig 19 that had much lower in the recovering group and consisted of one gene -- interferon-induced very large GTPase 1-like (GVINP1).
Interferons boost immune system processes and previous studies have linked interferon related genes to persisting post-Bd [@bryne2020] or increased expression when exposed to Bd in the lab [@rosenblum2008a; @ellison2014].
Variants in the specific gene identified here, GVINP1, explained \~20% of the resistance to Amoebic Gill Disease in Atlantic Salmon, a finding confirmed via expression analysis [@robledo2018; @robledo2020].
Furthermore, a key difference in susceptibility between populations of alpine tree frogs (*Litoria verreauxii alpina*) was the lower expression of IFN-induced very large GTPase in the more resistant, long-exposed population and upregulation of this same gene in more susceptible populations [@grogan2018].
Our study therefore contributes to the growing body of evidence linking this gene to disease resistance in general, and specifically to Bd survival in amphibians.
Finally, we found evidence of high FST and higher in recovering populations in the complement genes C6 and C7 <!--# It is clear why we talk about the other two regions (the ones highlighted in figure 2) and RIN3 - see comments above. But I guess the only reason this is interesting is because it is a complement region and it shows balancing rather than directional selection. If it feels like we can't justify discussing it we can cut it and just focus on the other two.  -->.
This could indicate that balancing selection is acting in this region of the genome, favoring a diverse set of alleles in the recovering group which has been shown for C6 in humans [@soejima2005].
Overall, key outlier genes in our dataset link directly to experimental and observational studies of immune responses to pathogens -- and often specifically to Bd responses in amphibians -- adding confidence to our conclusion that selection and rapid evolutionary change has contributed to survival in MYL frogs.

@@

-   <!--# Allie text -->One thing we could think about discussing is that we really don't know the effect of some of the genomic changes we describe (i.e., do the changes increase or decrease overall expression of that gene in response to Bd).
    For MYLF evidence suggests that they probably need a bit MORE of an immune response because of findings showing very little response in susceptible animals (Rosenblum et al. 2012) however many studies have found that immune overactivation is linked to susceptibility for Bd.
    This could be different for populations with different histories of Bd exposure.
    Perhaps Rana muscosa has a more recent history with Bd (maybe with another lineage) than for instance Atelopus.
    All of this has led to me think this whole question deserves another paper integrating the Atelopus and MYLF datasets, so maybe we save all this for that?
    Open to whatever you all think is relevant to this story.

@@

Raw reads were filtered for adapters and contaminants using fastp [@chen2018] and aligned to the *Rana muscosa* genome [NCBI SRA: SRS6533475, reference = @hon2020] with repetitive elements masked using bwa ("mem" mode) [@li2013].
Exact PCR Duplicates were marked using Picard.
Variants were then called following GATK best practices (v.4.2.0.0, @van2020). Briefly, raw variants were called for each sample using HaplotypeCaller and combined using CombineGVCFs.
Next, genotypes were jointly called using GenotypeGVCFs.
Variants were then hard filtered using gatk VariantFiltration using the following parameters to remove low-quality sites: QD \< 2.0, FS \> 40.0, SOR \> 3.0, MQ \< 50.0, MQRankSum \> 3.0, MQRankSum \< -3.0, ReadPosRankSum \> 3.0, ReadPosRankSum \< -3.0.
This initial filter resulted in 1,595,206 variant sites across 53 individuals.

@@

To visualize the genomic relationship of our populations we created a PCA using the glPCA function in the adegenet R package [@jombart2008].
Next, to detect regions of the genome under selection between naive and recovering populations, we used a multivariate linear mixed model to evaluate individual variants and a splined window analysis to evaluate larger genomic regions.
<!--# Improve wording in this and subsequent paragraph to clarify that the two approaches were used to identify regions under selection.  For example, for us genomics dummies it would be useful to see something like: "We used two related approaches to identify regions under selection. The first approach A, did X, Y, and Z but had C limitations.  The second approach B, did X, Y, and Z and D limitations."--> GEMMA calculates and incorporates a relatedness matrix for input samples, allowing us to account for relatedness and population structure when calculating likelihood ratio statistics for our association test.
For this analysis we used a stringent data filter to include only SNPs with \< 5% missing data (missing for no more than 2 individuals).
GEMMA then calculated the likelihood ratio statistic <!--# What is the null with this test (or what are the two models being compared with the LRT)?  Probably obvious for genomics folks, but not totally clear to me.  I am guessing we are testing whether the `population` coefficient (naive vs. persistent) is significant for describing variability in a SNP? --> for the set of 148,307 high quality SNPs across 127 contigs.
We then calculated two statistical thresholds for identifying outliers using a Bonferroni correction for a significance level of 0.05 (liberal set) and 0.01 (stringent set).
<!--# Could we say something like: "Outliers indicated that a SNP was differently represented in naive vs. persistent populations" or whatever is the correct language." --> Results were visualized using a Manhattan plot and qqplot.

@@
